
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(DropletUtils)
library(scran)
library(SpatialExperiment) #in terminal: sudo Rscript -e 'if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager", repos =                     "https://cran.rstudio.com"); BiocManager::install("SpatialExperiment")'
                                  #if this fails due to magick, first run:  sudo apt-get update
                                                                   # then: sudo apt-get install libmagick++-dev
                                                                   # then sudo R to open R session and install.packages("magick")     


projDir = "/mnt/fmlab/group_folders/lythgo02/Spatial/"
```

## after processing in 20240206 QC final workflow
#wont work if loading from scratch 
```{r}


 load_all_data <- "/run/user/1804238067/gvfs/sftp:host=clust1-sub-2,user=lythgo02/scratchc/fmlab/lythgo02/visium_data/logbatch/nnsvg"
# List files in the directory
files <- list.files(path = load_all_data, pattern = "^nnsvg", full.names = TRUE)

# Print out the list of files
print(files)

#assign names so they are loaded into the spe with the correct identifiers
logbatch <- c("SITSA3", "SITSB2", "SITSB4", 
                 "SITSC1","SITSC3","SITSD3",
                 "SITSE4","SITSF2", "SITSF4")

names(files) <- logbatch
# Initialize a list to store each sample
spe_list <- list()

# Loop through each file and process it
for (sample_name in names(files)) {
  # Read the Visium data
  spe <- readRDS(files[sample_name])
    # Add unique barcodes
  spe$barcode <- rownames(colData(spe))
  spe$barcodeid <- gsub("-1$", paste0("-", sample_name), spe$barcode)
  rownames(colData(spe)) <- spe$barcodeid
  spe$sample_id <- sample_name

  # Store the processed sample in the list
  spe_list[[sample_name]] <- spe 
  
  
  assign(paste0("spe_", sample_name), spe)
  #print progress check
  cat("Loaded:", sample_name, "\n")
}

```


## after processing in 20240206 QC final workflow
```{r}
 load_all_data <- "/run/user/1804238067/gvfs/sftp:host=clust1-sub-2,user=lythgo02/scratchc/fmlab/lythgo02/visium_data/arbitrary/nnsvg"
# List files in the directory
files <- list.files(path = load_all_data, pattern = "^nnsvg", full.names = TRUE)

# Print out the list of files
print(files)

#assign names so they are loaded into the spe with the correct identifiers
arbitrary <- c("SITSA1","SITSA2","SITSA4", "SITSB1", "SITSB3", 
                 "SITSC2","SITSC4","SITSD1","SITSD2", "SITSD4",
                 "SITSE2","SITSG2", "SITSH2")

names(files) <- arbitrary 

# Loop through each file and process it
for (sample_name in names(files)) {
 # Read the Visium data
  spe <- readRDS(files[sample_name])
    # Add unique barcodes
  #spe$barcode <- rownames(colData(spe))
 # spe$barcodeid <- gsub("-1$", paste0("-", sample_name), spe$barcode)
  #rownames(colData(spe)) <- spe$barcodeid
  #spe$sample_id <- sample_name
  
  # Store the processed sample in the list
  spe_list[[sample_name]] <- spe 
  
  
  assign(paste0("spe_", sample_name), spe)
  #print progress check
  cat("Loaded:", sample_name, "\n")
}

samplesheet <- cat(logbatch, arbitrary)


```

```{r}
#spe <- SEtools::mergeSEs(spe_list,commonOnly=FALSE, do.scale=FALSE) # cant' bind here because using objects prcoessed in scratch space

gene_count <- lapply(spe_list, function(x) table(rowData(x)$symbol))

gene_count_table <- do.call(cbind, gene_count) %>% as.data.frame() %>% 
  mutate(sum = rowSums(.)) %>% 
  filter(sum >22 | 
           sum < 22)


```
```{r}
# Step 1: Extract unique sample IDs from colData(spe)
sample_ids <- unique(colData(spe)$sample_id)

# Step 2: Initialize a list to store subsets and plots
spe_subset <- list()
plot_spots <- list()

# Step 3: Loop through each sample and generate plots
for (sample_id in sample_ids) {
  # Subset spe by sample_id
  spe_subset[[sample_id]] <- spe[, colData(spe)$sample_id == sample_id]
  
  # Generate and store the plot for this sample
  plot_spots[[sample_id]] <- plotSpots(spe_subset[[sample_id]])
}

# Step 4: Display the plots in a grid
# Example: Display the first 11 plots in a 2x4 grid
gridExtra::grid.arrange(grobs = plot_spots[1:11], ncol = 4)

# Example: Display plots from 12 to 22 in a 2x4 grid
gridExtra::grid.arrange(grobs = plot_spots[12:22], ncol = 4)

# Step 5: Faceting by sample_id
spots <- plotSpots(spe) +
  facet_wrap(~colData(spe)$sample_id)
print(spots)

```
```{r}
spe_afternnsvg<- readRDS("/run/user/1804238067/gvfs/sftp:host=clust1-sub-2,user=lythgo02/scratchc/fmlab/lythgo02/visium_data/logbatch/nnsvg/nnsvgprocessed_SITSA3.rds")

dec <- readRDS("/run/user/1804238067/gvfs/sftp:host=clust1-sub-2,user=lythgo02/scratchc/fmlab/lythgo02/visium_data/logbatch/decSITSA3.rds")
```



```{r feature selection}

#spe_filt <- read_rds("20241219_post_logNormCounts.rds")
# remove mitochondrial genes
#is_mito_filt <- grepl("(^MT-)|(^mt-)", rowData(spe_filt)$symbol)
#table(is_mito_filt)
#spe_filt <- spe_filt[!is_mito_filt, ]
#dim(spe_filt)
```


```{r}
readRDS("20250226_dec_arbitrary_and_log_mito.rds")

saveRDS(spe_filt, "20250226_post_modelgenevar_arb_and_log.rds")
```

```{r}
# select top HVGs to use in downstream analyses (eg PCA)
top_hvgs <- getTopHVGs(dec, prop = 0.1)
length(top_hvgs)

#visualise expression of top most variable genes with violin plot for each gene

plotExpression(spe_afternnsvg, features = top_hvgs[1:20], point_alpha = 0.05)
```
Note that HVGs are defined based only on molecular features (i.e. gene expression), and do not take any spatial information into account. 
If the biologically meaningful spatial information in this dataset mainly reflects spatial distributions of cell types, then relying on HVGs for downstream analyses may be sufficient. However, many datasets contain further spatial structure that is not captured in this way, which may be investigated using spatially-aware methods such as identifying spatially variable genes (SVGs).

Alternatively, we can apply methods to identify spatially variable genes (SVGs) instead of HVGs 
Here, we define SVGs as any genes with spatially correlated patterns of expression across the tissue area.

Several methods to identify SVGs in ST data have recently been developed:

nnSVG: available as an R package from Bioconductor and described by Weber et al. (2023)

SPARK-X: available as an R package from GitHub and described by Zhu, Sun, and Zhou (2021)

SPARK: available as an R package from GitHub and described by Sun, Zhu, and Zhou (2020)

SpatialDE: available as a Python package from GitHub and described by Svensson, Teichmann, and Stegle (2018)

Step 1: Run nnSVG on spatial transcriptomics data to identify spatially variable genes.
Step 2: After identifying these genes, integrate and align image data to:
Visualize gene expression patterns overlaid on histology/IF images.
Perform image-based segmentation or annotation to analyze gene expression in specific regions.
```{r}
#devtools::install_github("https://github.com/lmweber/nnSVG")

library(nnSVG)

# run nnSVG
set.seed(123)
#spe_nnSVG <- nnSVG(spe_filt) #i run this on the cluster via nnsvg_logbatch.r or nnsvg_arbitrary.r instead
spe_nnSVG<- read_rds(file = "/run/user/1804238067/gvfs/sftp:host=clust1-sub-2,user=lythgo02/scratchc/fmlab/lythgo02/visium_data/bulkprocess_until_nnsvg/nnsvg/20250227_spe_SITSA2.rds")
# show results
rowData(spe_nnSVG) #results stored in rowdata

```

The results are stored in the rowData of the SpatialExperiment object.
The main results of interest are:
LR_stat: likelihood ratio (LR) statistics used to rank SVGs
rank: rank of top SVGs according to LR statistics
pval: approximate p-values
padj: approximate p-values adjusted for multiple testing
prop_sv: effect size defined as proportion of spatial variance
```{r, Investigate results}
# number of significant SVGs
table(rowData(spe_nnSVG)$padj <= 0.05)
## 

#show results for top 10 SVGs

test <- rowData(spe_nnSVG)[order(rowData(spe_nnSVG)$rank)[1:500], ] %>% 
  as.data.frame()

# plot spatial expression of top-ranked SVG
ix <- which(rowData(spe_nnSVG)$rank == 1)
ix_name <- rowData(spe_nnSVG)$symbol[ix]
ix_name
```
```{r}
df <- as.data.frame(cbind(spatialCoords(spe_nnSVG), expr = counts(spe_nnSVG)[ix, ]))

df <- as.data.frame(cbind(
  spatialCoords(spe_nnSVG), 
  expr = as.vector(as.matrix(counts(spe_nnSVG))[ix, ])
))
ggplot(df, aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, 
               color = expr)) + 
  geom_point(size = 0.8) + 
  coord_fixed() + 
  scale_y_reverse() + 
  scale_color_gradient(low = "gray90", high = "blue", 
                       trans = "sqrt", breaks = range(df$expr), 
                       name = "counts") + 
  ggtitle(ix_name) + 
  theme_bw() + 
  theme(plot.title = element_text(face = "italic"), 
        panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank())

#ggsave("plots/20250214_nnsvg.png")
```

#various options from here:
#convert to zar object and use spatialdata to align images (20250214_python_sptial.py)
#continue in  with https://bookdown.org/sjcockell/ismb-tutorial-2023/practical-session-3.html#neighbour-graph-and-distance-matrix

for clustering (with histo classifiersas priors) etc (see spatial_workflow doc)




Dimensionality reduction: 
PCA uses a linear transformation of the data, which aims at defining new dimensions (axis) that capture most of the variance observed in the original data.
The results of PCA can be used in downstream analysis.

t-SNE and UMAP are both non-linear methods of dimensionality reduction. They aim at keeping similar cells together and dissimilar clusters of cells apart from each other.

SingleCellExperiment objects contain a slot that can store representations of our data in reduced dimensions. runPCA() function runs PCA on a SCE object, and returns an updated version of the single cell object with the PCA result added to the reducedDim slot.

```{r dimensionality reduction PCA}
# compute PCA
set.seed(123)
spe_filt <- runPCA(spe_filt, subset_row = top_hvgs) #restrict PCA to use only HVGs
#returns top 50 PCs by default
reducedDimNames(spe_filt)
dim(reducedDim(spe_filt, "PCA"))

percent.var <- attr(reducedDim(spe_filt), "percentVar")

#determine optimum number of PCs to use
chosen_elbow <- PCAtools::findElbowPoint(percent.var)

plot(percent.var, log = "y", xlab = "PC", ylab = "Variance explained (%)")
abline(v=chosen_elbow, col="dodgerblue")

```
#Run UMAP on the set of top 50 PCs and retain top 2 UMAP components for visualization 
```{r dim reduction UMAP}

# compute UMAP on top 50 PCs
set.seed(123)
spe_filt <- runUMAP(spe_filt, dimred = "PCA")

reducedDimNames(spe_filt)
dim(reducedDim(spe_filt, "UMAP"))

# update column names for easier plotting
colnames(reducedDim(spe_filt, "UMAP")) <- paste0("UMAP", 1:2)

# plot top 2 PCA dimensions
plotDimRed(spe_filt, type = "PCA")

```
```{r}
# plot top 2 UMAP dimensions
plotDimRed(spe_filt, type = "UMAP")
```

https://bioinformatics-core-shared-training.github.io/SingleCell_RNASeq_Jan24/UnivCambridge_ScRnaSeqIntro_Base/Markdowns/07_Dataset_Integration.html

```{r batch correction}

```


Cluster gene expression via scRNAseq methods
Graph-based clustering using the Walktrap method implemented in scran (Lun, McCarthy, and Marioni 2016), applied to the top 50 PCs calculated on the set of top HVGs.Assume that biologically informative spatial distribution patterns of cell types can be detected from the molecular features (gene expression).
```{r clustering to identify spatial domains}
# graph-based clustering
set.seed(123)
k <- 10
g <- buildSNNGraph(spe_filt, k = k, use.dimred = "PCA")
g_walk <- igraph::cluster_walktrap(g)
clus <- g_walk$membership
table(clus)

# store cluster labels in column 'label' in colData
colLabels(spe_filt) <- factor(clus)
```

```{r}
# plot clusters in spatial x-y coordinates
# Example of dynamically assigning colors based on the number of spots
# Example of a color palette generator
my_palette <- my_palette <- rainbow(37)


spots <- plotSpots(spe_filt, annotate = "label",
          palette = my_palette)

spots
```
```{r}
# plot clusters in PCA reduced dimensions
pca <- plotDimRed(spe_filt, type = "PCA", 
           annotate = "label", palette = my_palette)

pca
```
```{r}
# plot clusters in UMAP reduced dimensions
umap <- plotDimRed(spe_filt, type = "UMAP", 
           annotate = "label", palette = my_palette)

umap
```

```{r}
# set gene names as row names for easier plotting
rownames(spe_filt) <- rowData(spe_filt)$symbol

# test for marker genes
markers <- findMarkers(spe_filt, test = "binom", direction = "up")

# returns a list with one DataFrame per cluster
markers
```

```{r}
# plot log-fold changes for one cluster over all other clusters
# selecting cluster 1
interesting <- markers[[1]]
best_set <- interesting[interesting$Top <= 5, ]
logFCs <- getMarkerEffects(best_set)

pheatmap::pheatmap(logFCs, breaks = seq(-5, 5, length.out = 101))


```


Both the SpatialImage (SpI) and SpatialExperiment (SpE) class currently support two basic image transformations, namely, rotation (via rotateImg()) and mirroring (via mirrorImg()). Specifically, for a SpI/E x:

rotateImg(x, degrees) expects as degrees a single numeric in +/-[0,90,…,360].
Here, a (negative) positive value corresponds to (counter-)clockwise rotation.
mirrorImg(x, axis) expects as axis a character string specifying
whether to mirror horizontally ("h") or vertically ("v").
see https://www.bioconductor.org/packages/release/bioc/vignettes/SpatialExperiment/inst/doc/SpatialExperiment.html
